---
title: 'TMI - Homework L7'
author: 
  - Alfonzetti Giuseppe,
  - Castiglione Cristian, 
  - Iadevito Alessandro,
  - Pozza Francesco
header-includes:
   - \usepackage{amsmath}
output:
  pdf_document:
    toc: false
    number_sections: false
fontsize: 12pt
---

# Exercise 1:
Let weib.y be an i.i.d. sample from a gamma r.v.
```{r echo=FALSE}
weib.y = c(225, 171, 198, 189, 189, 135, 162, 135, 117, 162)
```

with density
$$\frac{1}{\Gamma(\theta)}3^\theta y^{\theta-1}e^{-3y}\quad y,\theta>0$$

1. Find the m.l.e. of $\theta$ solving the likelihood equation and also maximizing directly the log likelihood.
2. Plot the relative log likelihood and find the 0.90 Wald and deviance confidence intervals.
3. Repeat the previous points considering the parameterization $\omega=\log\theta$.
4. Assume a prior $N(0,\sigma_0^2)$ for $\omega$ and perform a Bayesian analysis of the data, using both analytical and
simulation-based solutions.

## Solution :

\newpage 
# Exercise 2:
Consider the following data
```{r}
data = c(1.434313, 1.122792, 1.205189, 0.08399036, 2.700203, 1.621289,
0.5877176, 1.337052, 4.893989, 2.386678, 2.631867, 0.2630924,
3.077384, 2.815827, 0.3646507)
```

as an i.i.d. sample from a gamma r.v. with shape $\alpha$ and scale $1/\beta$.

1. Find the m.l.e. and plot the relative log likelihood.
2. Repeat the previous point with the parameterization $(\alpha, \mu) = (\alpha, \alpha/\beta)$. Comment on the results.

## Solution:

\newpage
# Exercise 3:
Consider Exercise 7 pag. 158 of Davison (2003). Consider the model described in the exercise, with $Y_i$ , the i-th claim, distributed as an exponential with mean $\mu_i$ and with $\log \mu_i = \beta_0 + \beta_1 x_i$, where $x_i$ is equal to $0$ for claims of hospital A and $1$ for claims of hospital B.

1. Write the log likelihood for $(\beta_0 ,\beta_1)$ and make a contour plot.
2. Obtain numerically the maximum likelihood estimate and estimates of the standard errors for each component.
3. Write a function that computes the profile log likelihood for $\beta_1$ and plot it in a sensible range of parameter
values.
4. Compute a $0.95$ deviance confidence interval for $\beta_1$ and compare it with the corresponding Wald interval.
5. Use the profile log likelihood for $\beta_1$ to provide a p-value for the null hypothesis that the two hospitals have the
same expected claim value.
6. Perform a Bayesian analysis for $(\beta_0 ,\beta_1)$ assuming for the parameters independent normal priors with mean $0$
and standard deviation $10$.

## Solution:






